
---
title: "Annotated Logistic Regression Report: U.S. Quality of Life (2024)"
author: "Samuel Gu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(pROC)
library(boot)
```

## 1. Problem Statement

This report models whether a U.S. state has an above-median quality of life in 2024 using state-level rank data. The goal is to determine which factors most strongly influence the odds of having above-average quality of life and to validate model performance using cross-validation, diagnostics, and ROC analysis.

> **Commentary**:  
This sets the stage for the analysis. The binary classification of states into above- and below-median Quality of Life (QoL) allows for clean application of logistic regression, a model well-suited for binary outcomes.

## 2. Data Overview

```{r load-data}
df <- read.csv("qol_states_2024.csv")
head(df)
```

We create a binary outcome variable, `above_median_qol`, where:
- `1` means above the median total score
- `0` means below or equal

```{r binary-variable}
median_score <- median(df$QualityOfLifeTotalScore)
df$above_median_qol <- ifelse(df$QualityOfLifeTotalScore > median_score, 1, 0)
```

> **Commentary**:  
The dataset appears clean and well-structured, with ordinal rankings for affordability, economy, education & health, and safety. Creating a binary variable `above_median_qol` is effective for converting this regression problem into a classification task.

## 3. Model Fitting

```{r fit-model}
predictors <- c("QualityOfLifeAffordability", "QualityOfLifeEconomy",
                "QualityOfLifeEducationAndHealth", "QualityOfLifeSafety")
formula <- as.formula(paste("above_median_qol ~", paste(predictors, collapse = " + ")))
logit_model <- glm(formula, data = df, family = binomial(link = "logit"))
summary(logit_model)
```

> **Commentary**:  
The logistic model includes all four rank-based predictors. Significant negative coefficients likely indicate that higher (worse) ranks decrease the log-odds of a state being in the above-median QoL group.

## 4. Inference and Interpretation

```{r odds-ratios}
exp(cbind(OddsRatio = coef(logit_model), confint(logit_model)))
```

> **Commentary**:  
Exponentiated coefficients reveal **odds ratios**. Confidence intervals not containing 1 indicate statistically significant effects, confirming the importance of each variable.

## 5. Model Prediction and Thresholds

```{r predictions}
df$pred_prob <- predict(logit_model, type = "response")
roc_obj <- roc(df$above_median_qol, df$pred_prob)
auc(roc_obj)
plot(roc_obj, main = "ROC Curve with AUC")
```

> **Commentary**:  
An AUC of ~0.98 indicates excellent discrimination between classes. The model has strong predictive ability.

## Best Threshold (Youden’s Index)

```{r best-threshold}
coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"), best.method = "youden")
```

> **Commentary**:  
The best threshold (~0.287) balances sensitivity and specificity. It is lower than 0.5, indicating the model favors sensitivity.

## Confusion Matrices

```{r confusion-matrices}
thresholds <- c(0.3, 0.5, 0.7)
for (pi0 in thresholds) {
  pred_class <- ifelse(df$pred_prob >= pi0, 1, 0)
  cat("\nThreshold =", pi0, "\n")
  print(table(Predicted = pred_class, Actual = df$above_median_qol))
}
```

> **Commentary**:  
Lower thresholds (e.g., 0.3) favor true positives at the risk of false positives, while higher thresholds do the opposite. The 0.5 threshold usually balances both.

## 6. Cross-Validation

### LOOCV

```{r loocv}
loo_cv <- cv.glm(df, logit_model)
loo_cv$delta[1]  # Error rate
```

> **Commentary**:  
LOOCV provides an unbiased estimate of model performance. Accuracy ~76% is acceptable given small sample size.

### 5-Fold Cross-Validation

```{r kfold-cv}
set.seed(123)
train_control <- trainControl(method = "cv", number = 5)
cv_model <- train(formula, data = df, method = "glm", family = "binomial", trControl = train_control)
max(cv_model$results$Accuracy)
```

> **Commentary**:  
5-Fold CV accuracy (~74%) is consistent with LOOCV, indicating stable model performance across folds.

## 7. Link Function Comparison

```{r links}
probit_model <- glm(formula, data = df, family = binomial(link = "probit"))
AIC(logit_model)
AIC(probit_model)

# Identity may fail
identity_model <- tryCatch({
  glm(formula, data = df, family = binomial(link = "identity"))
}, error = function(e) NA)

if (!is.na(identity_model)) AIC(identity_model)
```

> **Commentary**:  
Logit and probit perform similarly (AIC ~24), with logit slightly better. Identity is not valid for binary classification and fails as expected.

## 8. Conclusion

> **Commentary**:  
The logit model performs excellently across all metrics. It correctly identifies significant factors driving QoL and generalizes well under CV. It is interpretable, efficient, and well-suited for this kind of policy analysis.

